Streaming $2$-Center with Outliers in High Dimension

### 2015-04-10 paper.tex
last revision on comments

@@

%--- from kim and ahn

In all the algorithms given in this section it is assumed that $p_1$ is a non-outlier point. It is not hard to see that, this limitation can be remedied by considering $\cO(z)$ parallel instances of the algorithm, similar to Algorithm~\ref{alg:1center}.
% Behnam: Move this to upper section, this assumption can be used for previous algorithm

@@

\subsection{The Case $\delta^* < C r*$}
\label{subsec:smaller}
Our idea in this section is derived from \cite[Sec.~3.1]{ahn2014computing, kim2014improved}. To avoid duplication, we just outline the important parts of their algorithm and our modifications to it. Kim's algorithm has $10$ different states. In each step one or many of the states can be valid and their algorithm considers all of them in parallel. In each state, we have two balls. A transition between the states occurs whenever a point not covered by any of the two balls has arrived in the stream. This transition graph starts with a blank node (with no balls) and has depth four (without considering the end nodes). It is straightforward to see that this transition graph is a DAG.
Our modification is on the transition part.
Points that are covered by current solution can safely be ignored, as they do not cause any transitions in the state, and only those points that lie outside the solution are candidates for being an outlier.
In each state the number of points that should be considered as outlier is unknown. So all the possible choices should be considered. It is sufficient to have four integers, $n_1,\dots, n_4$ representing the number of outliers in depth $1$ to $4$, such that $\sum_{i=1}^{4} n_i=z$. To sum up, the modified algorithm is defined as follows:
\begin{algorithm}
\label{alg:smallc}
\leavevmode
\begin{algorithmic}
\Function{Known$r'$}{$\mathcal{P}, z, r'$}
	\For{$(n_1, \dots, n_4) \gets \set{(n_1, \dots, n_4) \provided \sum n_i=z}$}
		\State $CS \gets B(p_1,r')$.
		\State $p_{o_2} \gets $ \Call{NthOutlier}{$n_1 + 1$, $p_1$}
		\If{$p_{o_2}$ arrives}
			\State $CS\gets$ the sub-case of $p_{o_2} \in B_{2, 1}^*$ $\cup$ the sub-case of $p_{o_2} \in B_{2,2}^*$.
			\For{each case $S$}
				\State $p_{o_3} \gets$ \Call{NthOutlier}{$n_2 + 1$, $p_{o_2}$}
				\If{$p_{o_3}$ arrives} 
					\State $CS\gets CS\setminus S$ $\cup$ the sub-case of $p_{o_3} \in B_{2, 1}^*$ $\cup$ the sub-case of $p_{o_3} \in B_{2,2}^*$.
					\For {each sub-case $S'$} 
						\State $p_{o_4} \gets$ \Call{NthOutlier}{$n_3 + 1$, $p_{o_3}$}
						\If {$p_{o_4}$ arrives} 
							\State Replace $S'$ with a new candidate solution.
							\State $p_{o_5} \gets$ \Call{NthOutlier}{$n_4 + 1$, $p_{o_4}$}
							\If{$p_{o_5}$ arrives}
								\State Abandon the solution.
							\EndIf
						\EndIf
					\EndFor
				\EndIf
			\EndFor

		\EndIf
		\State $\mathop{Sol}_{n_1,n_2,n_3,n_4} \gets$ the solution with the smallest larger radius, among all the sub-cases.
	\EndFor
	\State \Return the solution $\mathop{Sol}_{n_1,n_2,n_3,n_4}$ with the smallest radius, for all valid values of $(n_1,n_2,n_3,n_4)$.
\EndFunction
\end{algorithmic}
\end{algorithm}
Here, \textproc{NthOutlier}$(n,)$ returns the $n$\nth next point in $\mathcal{P}$ that does not lie in current candidate solution.

@@

\begin{lemma}
\label{lem:2lt1}
Let $r_1^*, r_2^*$ be the radius of $1$-center and $2$-center optimal solutions with $z$ outliers, respectively. Then $r_1^* < \left(\frac{C}{2} + 2\right) r_2^*$.
\end{lemma}
\begin{proof}
Suppose that $(B_{2, 1}^*(c_1^*, r_2^*), B_{2,2}^*(c_2^*, r_2^*))$ is an optimal solution for $2$-center problem. Consider $m$ as the midpoint of segment 
$c_1^*c_2^*$.
Clearly, $B\left(m, \frac{\delta}{2} + 2r_2^*\right)$ covers both $B_{2, 1}^*$ and $B_{2,2}^*$. So it is a solution for $1$-center problem and the proof is complete.
\end{proof}

@@

%In this section, we explain how we get r' such that 1.2r^*  r' < (1.2 + 2\eps/3)r^*. The basic procedure works as follows:

@@

Let $l_1=k_1$ and $l_i=2^j l_1$, where $l_i$ is the smallest number satisfying $l_i \ge k_i$. By Observation~\ref{obs:2apr}, $l_i$ is a $(2C + 8)$-approximation for $r_2^*$. If we divide the interval $(0, 1.2 l_i]$ into $m$ equal segments, then length of each segment is at most $2/3 \eps r_2^*$ and the endpoints as candidates are $\mathcal{L}_i = \set{j \times t_i \provided j = 1, \dots, m}$ where $t_i = \frac{1.2 l_i}{m}$. 
The inequality $1.2 r_2^* \le j \times l_i \le (1.2 + \frac{2\eps}{3})r_2^*$ holds for at least one of these candidates.

@@

%kiana: bacheha in paragraph paini kheili bad bud benazaram. man ye talashai rush kardam vali hanuzam benazaram jaye kar dare
%Behnam: khoobe
When the first non-zero radius is observed, $l_1$ is set. Algorithm~\ref{alg:smallc} should be executed for each candidate in $\mathcal{L}_1$ for all the previous points, which have been stored in a buffer. It is not hard to see that the space complexity for storing these points is $\cO(z)$. For each new point, if $l_i=l_{i-1}$ then $\mathcal{L}_i = \mathcal{L}_{i-1}$ and it suffices to add the new point to all parallel instances corresponding to candidates in $\mathcal{L}_i$. Otherwise, Algorithm~\ref{alg:smallc} should be executed for all of the candidates in $\mathcal{L}_i$, with all the points. But this is not achievable due to space limitations. To remedy
%kiana, sahand: remedy?? really??
this problem, note that if the candidate $x_j=j \times t_i \in \mathcal{L}_i$ is also in $\mathcal{L}_{i-1}$, there is no need to start Algorithm~\ref{alg:smallc} over and the current execution will be continued. If $x_j \notin \mathcal{L}_{i-1}$, then it is certain that $x_j \ge l_{i - 1}$ (Observe that $l_i=ul_{i-1}$, for some $u\in \mathbb{N}$). Since the ball $B_1(p_1, l_{i-1})$, obtained from Algorithm~\ref{alg:1center}, has at most $z$ outliers, all the non-outlier points of Algorithm~\ref{alg:1center}  lie in the candidate balls in Algorithm~\ref{alg:smallc}.
%kiana: in jomle balai oke be nazaretun? 
% Behnam: be nazare man gheire vazehe, be nazaram baiad begim hade aksar z ta biroonan, oonam oonaii ke too buffer algorithme 1 an
These outliers have been stored in a buffer, so for the new candidate length, it suffices to execute Algorithm~\ref{alg:smallc} with only the outlier points of $B_1$.

@@

%\begin{algorithmic}
%\Procedure{add to furthest buffer(z+1)}{p}
%	\State add p to buffer
%	\If{size of buffer $> z+1$}
%		\State remove nearest point from $p_1$
%	\EndIf
%\EndProcedure
%\end{algorithmic}

@@

Function \textproc{AddTo$B_2$} is used to add a point $p$ to $B_2$. Initially, $B_2$ is empty and $c_2$ and $r_c$ are not initialized. When an appropriate candidate is found, Algorithm~\ref{alg:case2} sets $c_2$ to that candidate and $r_c$ to $\frac{2\card{c_1c_2}}{C}$ and thereafter
%kiana: bacheha kudumetun in kalame ro gozashtin? azash motmaeni? man hichja nadide budamesh
points are added to $B_2$ only if they lie within $r_c$-radius of $c_2$. At this point $c_2$ is considered as the \emph{candidate point} $c_p$. Clearly the subdivision property for $B_2$ holds initially when $B_2$ consists of the single point $c_2$. All the subsequent points added to $B_2$ are within $r_c$-radius of $c_2$ and given $C\ge 7$, by Observation~\ref{obs:intersection} subdivision property is invariant under this operation.

@@

\begin{algorithm}
\label{alg:case2}
\leavevmode
\begin{algorithmic}
\State $\delta \gets 0, r_c \gets 0$
\State $c_1 \gets p_1$
\For{$p \in P$}
	\If{at least $z+1$ points have been processed}
		\State $q \gets$  $(z+1)$-furthest point from $c_1$
	\Else
		\State $q \gets c_1$
	\EndIf
	\State $\delta \gets \frac{C}{C+4} \card{c_1q}$
	\If{!\Call{AddTo$B_1$}{$p$} and !\Call{AddTo$B_2$}{$p$}}
		\State add $p$ to Buffer
		\While{$\card{\mbox{Buffer}} > z$}
			\If{$\card{B_2} \ge (d+1)(z+1)$}
				\State $B_1 \gets B_u$, $B_2 \gets \emptyset$
			\ElsIf{$c_2$ is set}
				\State $B_1 \gets B_1 \cup \set{c_2}$, $B_u \gets B_u \cup \set{c_2}$
			\EndIf
			\State $T \gets \mbox{Buffer} \cup B_2 - \set{c_2}$
			\State $\mbox{Buffer} \gets \emptyset, B_2 \gets \emptyset$
			\If{first iteration of while}
				\For{$p \in T$}
					\If{\Call{AddTo$B_1$}{$p$}}
						\State remove $p$ from $T$
					\EndIf
				\EndFor
			\EndIf
			\State $c_2 \gets$ arbitrary point in $T$
			\State $r_c \gets \frac{2}{C}\card{c_1 c_2}$
			\For{$p \in T$}
				\State \Call{AddTo$B_2$}{$p$}
			\EndFor				
			\State $\mbox{Buffer} \gets T \setminus B_2$
		\EndWhile
	\EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

Let $B_1'$, $B_2'$ and $B_u'$ be the $1$-centers obtained from $B_1$, $B_2$ and $B_u$ by the $\beta$-approximation algorithm, respectively. To answer a query, we use these balls. By the initial assumption about $p_1$, $B_1$ contains $B_{2,1}^*$ and $B_1'$ is a $\beta$-approximation for $B_{2,1}^*$. But for $B_2$ it may be the case that our assumption about the \emph{candidate point} was incorrect or non-optimal, so $B_2'$ is not a good approximation for $B_{2,2}^*$. There are two cases for $\card{B_2}$ to be considered. If $\card{B_2} < (z+1)(d+1)$, then we can try each point in $B_2 \cup \mbox{Buffer}$ as a candidate for the \emph{candidate point}. If $\card{B_2} \ge (d+1)(z+1)$, then there is no need for points in $B_2$ to be considered, since for those points as \emph{candidate point}, $B_1$ would be replaced by $B_u$ and $B_2$ would be empty. Thus it suffices to consider points of Buffer as candidates and compare them to the solution $(B_u, \emptyset)$.

\begin{algorithm}
\label{alg:query}
\leavevmode
\begin{algorithmic}
\Function{Query}{}
	\State solutions = [\Call{QuerySubdivision}{$B_1$ $B_2$ Buffer}]

@@

	\For{$c_2 \in$ candidates}
		\State $r_c \gets \frac{2}{C} \card{c_1 c_2}$
		\State $B_2 \gets \emptyset$, Buffer $\gets \emptyset$ 
		\For{$p \in$ candidates}
			\If{not \Call{AddTo$B_2$}{$p$}}
				\State add $p$ to Buffer
			\EndIf
		\EndFor
		\State add \Call{QuerySubdivision}{$B_1$, $B_2$, Buffer} to solutions
	\EndFor
	\State \Return \Call{$\min$}{solutions}
\EndFunction
\Statex
\Function{QuerySubdivision}{$B_1$, $B_2$, Buffer}
	\State solutions $\gets$ empty list
	\For{$k \gets 0, \dots,(z - \card{\mbox{Buffer}})$}
		\State $r \gets \max(1-\mbox{center}(B_1, k), 1-\mbox{center}(B_2, z - \card{\mbox{Buffer}} - k))$
		\State add $r$ to solutions
	\EndFor
	\State \Return \Call{$\min$}{solutions}
\EndFunction
\end{algorithmic}
\end{algorithm}

>>>>

In all the algorithms given in this section it is assumed that $p_1$ is a non-outlier point. It is not hard to see that, this limitation can be remedied by considering $\cO(z)$ parallel instances of the algorithm, similar to Algorithm~\ref{alg:1center}.
% Behnam: Move this to upper section, this assumption can be used for previous algorithm
%kiana: Here is better. Since we do not have p_1 in section 4.2

@@

@@

\subsection{The Case $\delta^* < C r*$}
\label{subsec:smaller}
Our idea in this section is derived from \cite[Sec.~3.1]{ahn2014computing, kim2014improved}. To avoid duplication, we just outline the important parts of their algorithm and our modifications to it. Kim's algorithm has $10$ different states. In each step one or many of the states can be valid and their algorithm considers all of them in parallel. In each state, we have two balls. A transition between the states occurs whenever a point not covered by any of the two balls has arrived in the stream. This transition graph starts with a blank node (with no balls) and has depth four (without considering the end nodes). It is straightforward to see that this transition graph is a DAG.
Our modification is on the transition part.
Points that are covered by current solution can safely be ignored, as they do not cause any transitions in the state, and only those points that lie outside the solution are candidates for being an outlier.
In each state the number of points that should be considered as outlier is unknown. So all the possible choices should be considered. It is sufficient to have four integers, $n_1,\dots, n_4$ representing the number of outliers in depth $1$ to $4$, such that $\sum_{i=1}^{4} n_i=z$. To sum up, the modified algorithm is defined as follows:
{
\algrenewcommand\algorithmicindent{0.8em}%
\begin{algorithm}
\label{alg:smallc}
\leavevmode
\begin{algorithmic}
\Function{Known$r'$}{$\mathcal{P}, z, r'$}
	\For{$(n_1, \dots, n_4) \gets \set{(n_1, \dots, n_4) \provided \sum n_i=z}$}
		\State $CS \gets B(p_1,r')$.
		\State $p_{o_2} \gets $ \Call{NthOutlier}{$n_1 + 1$, $p_1$}
		\If{$p_{o_2}$ arrives}
			\State $CS\gets$ the sub-case of $p_{o_2} \in B_{2, 1}^*$ $\cup$ the sub-case of $p_{o_2} \in B_{2,2}^*$.
			\For{each case $S$}
				\State $p_{o_3} \gets$ \Call{NthOutlier}{$n_2 + 1$, $p_{o_2}$}
				\If{$p_{o_3}$ arrives} 
					\State $CS\gets CS\setminus S$ $\cup$ the sub-case of $p_{o_3} \in B_{2, 1}^*$ $\cup$ the sub-case of $p_{o_3} \in B_{2,2}^*$.
					\For {each sub-case $S'$} 
						\State $p_{o_4} \gets$ \Call{NthOutlier}{$n_3 + 1$, $p_{o_3}$}
						\If {$p_{o_4}$ arrives} 
							\State Replace $S'$ with a new candidate solution.
							\State $p_{o_5} \gets$ \Call{NthOutlier}{$n_4 + 1$, $p_{o_4}$}
							\If{$p_{o_5}$ arrives}
								\State Abandon the solution.
							\EndIf
						\EndIf
					\EndFor
				\EndIf
			\EndFor

		\EndIf
		\State $\mathop{Sol}_{n_1,n_2,n_3,n_4} \gets$ the solution with the smallest larger radius, among all the sub-cases.
	\EndFor
	\State \Return the solution $\mathop{Sol}_{n_1,n_2,n_3,n_4}$ with the smallest radius, for all valid values of $(n_1,n_2,n_3,n_4)$.
\EndFunction
\end{algorithmic}
\end{algorithm}
}
Here, \textproc{NthOutlier}$(n,p)$ returns the $n$\nth point arriving after $p$ in $\mathcal{P}$ that does not lie in current candidate solution.

@@

\begin{lemma}
\label{lem:2lt1}
Let $r_1^*, r_2^*$ be the radius of $1$-center and $2$-center optimal solutions with $z$ outliers, respectively. Then $r_1^* < \left(\frac{C}{2} + 2\right) r_2^*$.
\end{lemma}
\begin{proof}
Suppose that $(B_{2, 1}^*(c_1^*, r_2^*), B_{2,2}^*(c_2^*, r_2^*))$ is an optimal solution for $2$-center problem. Consider $m$ as the midpoint of segment 
$c_1^*c_2^*$ (Figure~\ref{fig:2lt1}).
Clearly, $B\left(m, \frac{\delta}{2} + 2r_2^*\right)$ covers both $B_{2, 1}^*$ and $B_{2,2}^*$. So it is a solution for $1$-center problem and the proof is complete.
\end{proof}

@@

@@

Let $l_1=k_1$ and $l_i=2^j l_1$, where $l_i$ is the smallest number satisfying $l_i \ge k_i$. By Observation~\ref{obs:2apr}, $l_i$ is a $(2C + 8)$-approximation for $r_2^*$. If we divide the interval $(0, 1.2 l_i]$ into $m$ equal segments (Figure~\ref{fig:findr}), then length of each segment is at most $2/3 \eps r_2^*$ and the endpoints as candidates are $\mathcal{L}_i = \set{j \times t_i \provided j = 1, \dots, m}$ where $t_i = \frac{1.2 l_i}{m}$. 
The inequality $1.2 r_2^* \le j \times l_i \le (1.2 + \frac{2\eps}{3})r_2^*$ holds for at least one of these candidates.

@@

When the first non-zero radius is observed, $l_1$ is set. Algorithm~\ref{alg:smallc} should be executed for each candidate in $\mathcal{L}_1$ for all the previous points, which have been stored in a buffer. It is not hard to see that the space complexity for storing these points is $\cO(z)$. For each new point, if $l_i=l_{i-1}$ then $\mathcal{L}_i = \mathcal{L}_{i-1}$ and it suffices to add the new point to all parallel instances corresponding to candidates in $\mathcal{L}_i$. Otherwise, Algorithm~\ref{alg:smallc} should be executed for all of the candidates in $\mathcal{L}_i$, with all the points. But this is not achievable due to space limitations. To remedy this problem, note that if the candidate $x_j=j \times t_i \in \mathcal{L}_i$ is also in $\mathcal{L}_{i-1}$, there is no need to start Algorithm~\ref{alg:smallc} over and the current execution will be continued. If $x_j \notin \mathcal{L}_{i-1}$, then it is certain that $x_j \ge l_{i - 1}$ (Observe that $l_i=ul_{i-1}$, for some $u\in \mathbb{N}$). Since the points that do not lie in the candidate solution are saved in the buffer used in Algorithm~\ref{alg:1center} (which has size at most $z$), all the non-outlier points of this algorithm lie in the candidate balls in Algorithm~\ref{alg:smallc}.
%kiana: added the above sentece instead of the below
%Since the ball $B_1(p_1, l_{i-1})$, obtained from Algorithm~\ref{alg:1center}, has at most $z$ outliers, all the non-outlier points of Algorithm~\ref{alg:1center}  lie in the candidate balls in Algorithm~\ref{alg:smallc}.
These outliers have been stored in a buffer, so for the new candidate length, it suffices to execute Algorithm~\ref{alg:smallc} with only the outlier points of $B_1$.

@@

@@

Function \textproc{AddTo$B_2$} is used to add a point $p$ to $B_2$. Initially, $B_2$ is empty and $c_2$ and $r_c$ are not initialized. When an appropriate candidate is found, Algorithm~\ref{alg:case2} sets $c_2$ to that candidate and $r_c$ to $\frac{2\card{c_1c_2}}{C}$ and thereafter points are added to $B_2$ only if they lie within $r_c$-radius of $c_2$. At this point $c_2$ is considered as the \emph{candidate point} $c_p$. Clearly the subdivision property for $B_2$ holds initially when $B_2$ consists of the single point $c_2$. All the subsequent points added to $B_2$ are within $r_c$-radius of $c_2$ and given $C\ge 7$, by Observation~\ref{obs:intersection} subdivision property is invariant under this operation.

@@

{
\algrenewcommand\algorithmicindent{0.8em}%
\begin{algorithm}
\label{alg:case2}
\leavevmode
\begin{algorithmic}
\State $\delta \gets 0, r_c \gets 0$
\State $c_1 \gets p_1$
\For{$p \in P$}
	\If{at least $z+1$ points have been processed}
		\State $q \gets$  $(z+1)$-furthest point from $c_1$
	\Else
		\State $q \gets c_1$
	\EndIf
	\State $\delta \gets \frac{C}{C+4} \card{c_1q}$
	\If{!\Call{AddTo$B_1$}{$p$} and !\Call{AddTo$B_2$}{$p$}}
		\State add $p$ to Buffer
		\While{$\card{\mbox{Buffer}} > z$}
			\If{$\card{B_2} \ge (d+1)(z+1)$}
				\State $B_1 \gets B_u$, $B_2 \gets \emptyset$
			\ElsIf{$c_2$ is set}
				\State $B_1 \gets B_1 \cup \set{c_2}$, $B_u \gets B_u \cup \set{c_2}$
			\EndIf
			\State $T \gets \mbox{Buffer} \cup B_2 - \set{c_2}$
			\State $\mbox{Buffer} \gets \emptyset, B_2 \gets \emptyset$
			\If{first iteration of while}
				\For{$p \in T$}
					\If{\Call{AddTo$B_1$}{$p$}}
						\State remove $p$ from $T$
					\EndIf
				\EndFor
			\EndIf
			\State $c_2 \gets$ arbitrary point in $T$
			\State $r_c \gets \frac{2}{C}\card{c_1 c_2}$
			\For{$p \in T$}
				\State \Call{AddTo$B_2$}{$p$}
			\EndFor				
			\State $\mbox{Buffer} \gets T \setminus B_2$
		\EndWhile
	\EndIf
\EndFor
\end{algorithmic}
\end{algorithm}
}

Let $B_1'$, $B_2'$ and $B_u'$ be the $1$-centers obtained from $B_1$, $B_2$ and $B_u$ by the $\beta$-approximation algorithm, respectively. To answer a query, we use these balls. By the initial assumption about $p_1$, $B_1$ contains $B_{2,1}^*$ and $B_1'$ is a $\beta$-approximation for $B_{2,1}^*$. But for $B_2$ it may be the case that our assumption about the \emph{candidate point} was incorrect or non-optimal, so $B_2'$ is not a good approximation for $B_{2,2}^*$. There are two cases for $\card{B_2}$ to be considered. If $\card{B_2} < (z+1)(d+1)$, then we can try each point in $B_2 \cup \mbox{Buffer}$ as a candidate for the \emph{candidate point}. If $\card{B_2} \ge (d+1)(z+1)$, then there is no need for points in $B_2$ to be considered, since for those points as \emph{candidate point}, $B_1$ would be replaced by $B_u$ and $B_2$ would be empty. Thus it suffices to consider points of Buffer as candidates and compare them to the solution $(B_u, \emptyset)$.
{
\algrenewcommand\algorithmicindent{0.8em}%
\begin{algorithm}
\label{alg:query}
\leavevmode
\begin{algorithmic}
\Function{Query}{}
	\State solutions = [\Call{QuerySubdivision}{$B_1$ $B_2$ Buffer}]

@@

	\For{$c_2 \in$ candidates}
		\State $r_c \gets \frac{2}{C} \card{c_1 c_2}$
		\State $B_2 \gets \emptyset$, Buffer $\gets \emptyset$ 
		\For{$p \in$ candidates}
			\If{not \Call{AddTo$B_2$}{$p$}}
				\State add $p$ to Buffer
			\EndIf
		\EndFor
		\State add \Call{QuerySubdivision}{$B_1$, $B_2$, Buffer} to solutions
	\EndFor
	\State \Return \Call{$\min$}{solutions}
\EndFunction
\Statex
\Function{QuerySubdivision}{$B_1$, $B_2$, Buffer}
	\State solutions $\gets$ empty list
	\For{$k \gets 0, \dots,(z - \card{\mbox{Buffer}})$}
		\State $r \gets \max(1-\mbox{center}(B_1, k), 1-\mbox{center}(B_2, z - \card{\mbox{Buffer}} - k))$
		\State add $r$ to solutions
	\EndFor
	\State \Return \Call{$\min$}{solutions}
\EndFunction
\end{algorithmic}
\end{algorithm}
}
